---
title: "EDA Final Project"
author: "Dov Tuch"
output: 
  pdf_document: default
  html_notebook: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = F}
library(tidyverse)
library(tidytext)
library(caret)
library(patchwork)
library(ggrepel)
```

```{r, include= FALSE}
setwd("~/Documents/Studies/Ds_applications/Final_ProjectTuchdov")
food_train <- read_csv("data/food_train.csv")
nutrients <- read_csv("data/nutrients.csv")
food_nutrients <- read_csv("data/food_nutrients.csv")


```

# Welcome

Hello Giora, This EDA can be summed up into 2 parts. The first one is
just general EDA with no story or purpose behind it. The second part is
a market analysis I conducted with the data at hand, with a mission and
a little bit of sarcasm :) Hope you enjoy (the second part).

# First part: preparing the data set and general EDA.

I will make the 3 data sets to one single tibble. Where the columns are
the nutritional values, and the first 10 ingredients in each snack.

```{r, warning= FALSE, results='hide'}
fst_10ings = str_replace_all(food_train$ingredients, "[().*+-]", "") %>% 
str_split_fixed( ',',n = 11) %>% as_tibble() %>% 
  select(-V11)

```

```{r, echo= F, results='hide',include=FALSE}
# import the dataset where the nuritional values are columns, the code for this 
# data frame can be located at fp_model2
food_nut = read_csv('/home/dovikde/Documents/Studies/Ds_applications/Final_ProjectTuchdov/all_nuts.csv')

```

```{r, results='hide'}
# removing zero variance columns
zv_cols = nearZeroVar(food_nut, names = T)

food_nut2 = food_nut %>%
  select(-all_of(zv_cols))
food_joined = food_train %>% 
  left_join(food_nut2, by = 'idx') 
food_complete = bind_cols(food_joined,fst_10ings )

```

```{r}
food_train %>% 
  distinct(serving_size_unit)
```

```{r}
food_complete %>% 
  filter(serving_size_unit == 'ml') %>% 
  nrow()
```

Because we have only 8 observations that are in ml units (out of \~
32K), I will drop the *serving_size_unit* column and those observations.
The data set that I will be working with will look like where v_i is the
i-th appearing ingredient in the ingredients column:

```{r, echo=FALSE}
food_final = food_complete %>%
  filter(serving_size_unit == 'g') %>% 
  select(-c(serving_size_unit, ingredients,...9))
food_final <- food_final %>%
   mutate(across(everything(), ~ifelse(.=="", NA, as.character(.))),
          across(c(serving_size,'1003':'2000'), ~ as.numeric(.))) %>% 
  rename(Protein = '1003', Total_fat = '1004', Carbohydrate = '1005',
         Energy = '1008', Total_fiber = '1079', Calcium = '1087', Sodium ='1093',
         Iron = '1089',Saturated_fat = '1258', Sugars ='2000') %>% 
  rename_with(str_to_lower)
food_final %>% 
  glimpse()

```

## Correlations between the nutrients across all samples and by categories.

```{r,results='hide'}
# helper functions
# Get lower diagonal of the correlation matrix
get_lower_diag<-function(cormat){
    cormat[lower.tri(cormat)] <- NA
    return(cormat)
  }
reorder_cormat <- function(cormat){
# Use correlation between variables as distance
dd <- as.dist((1-cormat)/2)
hc <- hclust(dd)
cormat <-cormat[hc$order, hc$order]
}

# correltaion function heatmap
corr_mat = function(category_data){
  var_name = paste0(category_data)
  cor_nut = food_final %>% 
  filter(category == var_name) %>% 
  select(protein:sugars) %>% 
  cor() %>% 
  round(2)
  titleplot = str_replace_all(var_name, '_', " ")
  
  lower_diag= get_lower_diag(cor_nut)
  melted_cormat= reshape2::melt(lower_diag)
  heatmap_fun = ggplot(data = melted_cormat, aes(Var2, Var1, fill = value))+
   geom_tile(color = "ghostwhite")+
   scale_fill_gradient2(low = "blue1", high = "tomato2", mid = "ghostwhite", 
     midpoint = 0, limit = c(-1,1), space = "Lab", 
     ) +
    ggtitle(titleplot)+
    theme_minimal()+ 
   # theme(axis.text.x = element_text(angle = 45, vjust = 1, 
   #    size = 12, hjust = 1))+
   coord_fixed()+
    geom_text(aes(Var2, Var1, label = value), color = "black", size = 2)+ 
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.grid.major = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.ticks = element_blank(),
    axis.text.x = element_text(angle = 45, vjust = 1, 
    size = 12, hjust = 1))
  
  return(heatmap_fun)
  
}
```

```{r, echo = F}
cor_nut3 = food_final %>% 
  select(protein:sugars) %>% 
  cor() %>% 
  round(2)
  
  lower_diag= get_lower_diag(cor_nut3)
  melted_cormat= reshape2::melt(lower_diag)
  my_heatmap = ggplot(data = melted_cormat, aes(Var2, Var1, fill = value))+
   geom_tile(color = "ghostwhite")+
   scale_fill_gradient2(low = "blue1", high = "tomato2", mid = "ghostwhite", 
     midpoint = 0, limit = c(-1,1), space = "Lab", 
     ) +
    ggtitle("correlation across all samples")+
    theme_minimal()+ 
   # theme(axis.text.x = element_text(angle = 45, vjust = 1, 
   #    size = 12, hjust = 1))+
   coord_fixed()+
    geom_text(aes(Var2, Var1, label = value), color = "black", size = 1)+ 
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.grid.major = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.ticks = element_blank(),
    axis.text.x = element_text(angle = 45, vjust = 1, 
    size = 12, hjust = 1))
  
  my_heatmap
  
```

For all the snacks. total fat and protein are strongly inversely
correlated. With carbohydrates. While energy and total fat are highly
correlated.

```{r,echo=FALSE}

cor_choc = corr_mat('chocolate')
cor_cookies = corr_mat('cookies_biscuits')
cor_cakes = corr_mat('cakes_cupcakes_snack_cakes')
cor_popcorn = corr_mat('popcorn_peanuts_seeds_related_snacks')
cor_candy = corr_mat('candy')
cor_chips = corr_mat("chips_pretzels_snacks")


cor_choc + cor_cookies + plot_layout(guides = 'collect') +
  plot_annotation(title = 'The corelation of nutrients by category')
cor_cakes + cor_popcorn + plot_layout(guides = 'collect')
cor_candy + cor_chips  + plot_layout(guides = 'collect')

```

When we look at the nutrient correlation by category we find different
results. Candies total fat is very strongly correlated with saturated
fat, but for peanuts There is no correlation between the two nutritional
values. In the cakes category carbohydrates are strongly correlated with
sugars but not with chips. This realtionships can be used later as
predictors for each category.

```{r, echo= F, results='hide',message= FALSE}

fd_final = food_final %>% 
  mutate(across((v1:v10), ~ str_replace_all(.,"[^[\\da-zA-Z ]]","")),
         across((v1:v10), ~ str_replace_na(.)))

ing1 = fd_final %>% 
  group_by(category, v1) %>% 
  summarise(n = n()) %>% 
  mutate(nut_prop = n /sum(n),v1) %>% 
  arrange(desc(nut_prop))

ing2 = fd_final %>% 
  group_by(category, v2) %>% 
  summarise(m = n()) %>% 
  mutate(nut_prop2 = m /sum(m)) %>% 
  select(nut_prop2, category, v2) %>% 
  arrange(desc(nut_prop2))

ing3 = fd_final %>% 
  group_by(category, v3) %>% 
  summarise(m = n()) %>% 
  mutate(nut_prop3 = m /sum(m)) %>% 
  select(nut_prop3, category, v3) %>% 
  arrange(desc(nut_prop3))

ing1_top_5 = ing1 %>% 
  group_by(category) %>% 
  slice_max(order_by = nut_prop, n =1 )

ing2_top_5 = ing2 %>% 
  group_by(category) %>% 
  slice_max(order_by = nut_prop2, n =1 )

ing3_top_5 = ing3 %>% 
  group_by(category) %>% 
  slice_max(order_by = nut_prop3, n =1 )
```

```{r, echo= F}
table_0 = ing1_top_5 %>% 
  bind_cols(ing2_top_5 %>% 
              ungroup() %>% 
              select(-category)) %>% 
  bind_cols(ing3_top_5 %>% 
              ungroup() %>% 
              select(-category)
              ) %>% 
  select(-n)
```

```{r}
table_0 = table_0 %>% 
  mutate(category = 
  case_when(category == "cakes_cupcakes_snack_cakes" ~ 'cakes',
            category =="chips_pretzels_snacks" ~ 'chips',
            category == 'cookies_biscuits' ~ 'cookies',
            category == 'popcorn_peanuts_seeds_related_snacks' ~ 'snack related',
            TRUE ~ as.character(category))
  )
```

Let's look at the most occurring ingredients by category.

```{r, echo= F}

table_0 %>% 
  ggplot(aes(x = category, y = nut_prop, fill = v1))+
  geom_bar(stat = "identity")+
  plot_annotation(title = 'The proprtion of the top most used ingredient by category')+
  ylab('most common ingredient')+
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, 
    size = 8, hjust = 1))+
  guides(fill= guide_legend(title = "ingredient"))

table_0 %>% 
  ggplot(aes(x = category, y = nut_prop2, fill = v2))+
  geom_bar(stat = "identity")+
  plot_annotation(title = 'The proprtion of the top 2nd most used ingredient by category')+
  ylab('most common ingredient')+
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, 
    size = 8, hjust = 1))+
  guides(fill= guide_legend(title = "ingredient"))

table_0 %>% 
  ggplot(aes(x = category, y = nut_prop3, fill = v3))+
  geom_bar(stat = "identity")+
  ylab('most common ingredient')+
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, 
    size = 8, hjust = 1)
   )+
  guides(fill= guide_legend(title = "ingredient"))

```

We can get a lot of predicting power with those values, when potatoes
are the first ingredient we can have a good guess that it's from the
chips category. It's not a big surprise that peanuts and seeds have
mostly one ingredient, that's why the 2nd and 3rd most common
"ingredient" is NA.

# Second Part: Market Analysis!

During this EDA my mother started to nag me that I don't do anything
with my Econ major, just statistics all day. So to make mummy proud I
decided to go from general EDA to a market research!

# First question, where is the competition?

To answer this we need to understand how many brands are in each
category? To measure this we will look at the ratio of
n_brands/n_products in the category. We will call this measure as the
*Diversity*. so if the *Diversity* is high there are many brands in this
market and the highest value it can get is 1. When there is low market
diversity it means that there are less players in the market and so the
diversity will be close to 0.

```{r, echo = F}
food_train %>% 
  count(category) %>% 
  left_join(
    food_train %>% 
    count(category,brand) %>% 
    group_by(category) %>% 
    count(category) %>% 
    rename(num_brands = n),
    by = 'category'
    )
```

```{r, include = F}
table1 <- food_train %>% 
  count(category,brand) %>% 
  group_by(category) %>% 
  count(category) %>% 
  rename(num_brnds = n) %>% 
  left_join(food_train %>%  count(category),
            by = "category") %>% 
  mutate(snacks_brnd =  n/ num_brnds,
         # diversity 0 means monoply 1 means big competition
         diversity = num_brnds / n) %>% 
  rename(total_snacks = n) %>% 
  mutate(category = 
  case_when(category == "cakes_cupcakes_snack_cakes" ~ 'cakes',
            category =="chips_pretzels_snacks" ~ 'chips',
            category == 'cookies_biscuits' ~ 'cookies',
            category == 'popcorn_peanuts_seeds_related_snacks' ~ 'snack related',
            TRUE ~ as.character(category))
  )
```

```{r}
table1 %>% 
  ggplot(mapping = aes(x = category, y = diversity, fill = category))+
  geom_col()+
  theme(
        axis.ticks.x=element_blank(),
        legend.position  = "none"
        )+
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, 
    size = 11, hjust = 1)
   )+
  #geom_text(aes( label = paste0("n = ", total_snacks)), nudge_y = +0.02)+
  geom_label(aes(label = round(diversity, 2)))
```

We can see that popcorn peanuts and snacks have the lowest diversity.
But it also the most common category in the training sample. As always,
this breaks some economical models assumptions. As we saw before a large
parcentege of those products have only one ingredient (nuts or seeds).
It's a generic product so we expect it to have more sellers. Another
possible explanation is that most of these sellers operate on a small
business scale, So They don't need an FDA registration to sell their
nuts.

This leads as to the second question..

# Question2: Why there is less competition in the popcorn market?

Let's dig further into popcorn category, I will look at the brands and
look if there are any significant players.

```{r}
set.seed(12323)
food_train %>%
  filter(category == "popcorn_peanuts_seeds_related_snacks" ) %>%
  count(brand, sort = TRUE) %>% 
  mutate(y_random = c(runif(1265, min = -1, max = 1))) %>% 
  filter(n > 50) %>% 
  ggplot(aes(x = n, y = y_random, label = brand, color = as.factor(n))) +
  ggtitle('Brands that sell more then 50 items in the seed category')+
  geom_text_repel(max.overlaps = 1265) +
  theme_minimal() +
  theme(legend.position = "None",
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +  
  labs(x = "Offered products",
       y = "Brand name")
```

Looking at the biggest sellers here (meijer and target) we can
understand that the biggest players are not food focused companies but
big retailers. with much less products pharmacies like cvs and walgreens
are also big players. only nabisco is a company that focuses mainly on
packaged foods.

So I saw that retail stores and pharmacies of the branded snacks in the
seeds category. Let's see what is the situation in a category with more
*diversity*. for example the cookies category.

```{r}
set.seed(12323)
food_train %>%
  filter(category == "cookies_biscuits" ) %>%
  count(brand, sort = TRUE) %>% 
  mutate(y_random = c(runif(1274, min = -1, max = 1))) %>% 
  filter(n > 50) %>% 
  ggplot(aes(x = n, y = y_random, label = brand, color = as.factor(n))) +
  ggtitle('Brands that sell more then 50 items in the coockie category')+
  geom_text_repel(max.overlaps = 1274) +
  theme_minimal() +
  theme(legend.position = "None",
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +  
  labs(x = "Offered products",
       y = "Brand name")
```

In the cookie market only two out of the largest 5 selling brands are
not food focused. Nabisco biscuit company lives up to it's name :). To
support the effectiveness I gave of measuring the competition we can see
here that there are less brands that sell more then 50 products. (12 in
this category and 29 in the seed category) although there are slightly
more brands in this category then the last one (1274 and 1265)/

the only category where the top 5 sellers are food focused companies is
the chocolate category. To be in the chocolate business you really need
to focus.

# Question 3: how is nabisco different?

let's analyze nabisco in the cookie category,how are they're product
differ from the the others?

I will conduct a paired t test for each nutritional value with the
hypothesis that the means of nabisco are the same as the avarege cookie
& biscuit company.

```{r}
food_final %>% 
  filter(category == "cookies_biscuits") %>% 
  select(-c(idx, category)) %>% 
  summarise(across(where(is.numeric), list(avg = mean, sd = sd))) %>% 
  bind_rows(food_final %>% 
  filter(category == "cookies_biscuits") %>% 
  filter(brand == "nabisco biscuit company") %>% 
  select(-c(idx, category)) %>% 
  summarise(across(where(is.numeric), list(avg = mean, sd = sd))))
  
```

```{r}
cookie_avg = food_final %>% 
  filter(category == "cookies_biscuits") %>% 
  select(-c(idx, category)) %>% 
  summarise(across(where(is.numeric), list(avg = mean))) %>% 
  slice(1) %>% 
  as.numeric()

cookie_sd =food_final %>% 
  filter(category == "cookies_biscuits") %>% 
  select(-c(idx, category)) %>% 
  summarise(across(where(is.numeric), list(sd = sd))) %>% 
  slice(1) %>% 
  as.numeric()

nobosco_avgs = food_final %>% 
  filter(category == "cookies_biscuits") %>% 
  filter(brand == "nabisco biscuit company") %>% 
  select(-c(idx, category)) %>% 
  summarise(across(where(is.numeric), list(avg = mean))) %>% 
  slice(1) %>% 
  as.numeric()

nobosco_sd = food_final %>% 
  filter(category == "cookies_biscuits") %>% 
  filter(brand == "nabisco biscuit company") %>% 
  select(-c(idx, category)) %>% 
  summarise(across(where(is.numeric), list(sd = sd))) %>% 
  slice(1) %>% 
  as.numeric()
```

```{r}
all_ttest = function(nb_avg, nb_sd, all_avg, all_sd,n1 ,n2){
  m = length(nb_avg)
  t = numeric(m)
  for (i in 1:m){
    sp = sqrt(((n1- 1)*(nb_sd[i])^2 + (n2-1)*(all_sd[i])^2)/(n1+n2 -2))
    t[i] = abs((nb_avg[i]-all_avg[i])/(sp*sqrt(1/n1+1/n2))) > qt(0.975,n1 + n2- 2)
  }
  return(t)
}



b = all_ttest(nobosco_avgs,nobosco_sd,cookie_avg,cookie_sd, 140,5284)

```

From this test I can search for nad reject the null hypothesis with 0.05
significance that carbohydrate, energy, calcium, sodium and saturated
fats are different in the nobisco brand from the average cookie/biscuit
product.

And That's It. Good luck with the rest of your life!

Thank you for all the mind blows I had in this semester in your course.
